A data-driven model of nucleosynthesis with chemical tagging in a lower-dimensional latent space

### 2019-09-17 article/reviewer-response-v0.txt

4.  REVIEWER:

    Section 2 : 
    It would be highly beneficial for section 2 to potentially have a schematic (figure) of the method the authors outline. Especially in terms of the structure of how the code might be set up.

    RESPONSE:

    We have included a schematic to try to help visualise the method (new Figure 1).

    We have also added an Appendix that references the online documentation, and includes a code example. This code will generate fake data, fit those data, and produce some figures.

@@

9.  REVIEWER:

    Section 3.1 :
    The authors build a toy model in which they sample 100,000 'stars' using 20 clusters in a 5-D factor load space. The toy model seems to work (as per Fig.1). However when they move on to real data in section 3.3 they first only use a sample size of 1,000 (1% the sample size in the toy model) stars. It would be good for the authors to recompute their toy model sampling N=100,000 and N=1,000 'stars' and add this to Figure 1 to better illustrate the effect of lowering the sample size.

    RESPONSE:

    We have updated Figure 1 to show the grid search using N = 100,000 'stars' and N = 5,000 'stars'. We have also added the following paragraphs to Section 2:

    "We repeated this toy model experiment using a smaller sample size ($N = 5,000$) to be more representative of the sample sizes in later \Galah\ experiments (Section~\ref{sec:exp4}). The results of the grid search are also shown in Figure~\ref{fig:experiment-1-gridsearch}. Here BIC estimates the true number of latent factors correctly, but tends to underestimate the true number of clusters, more so than the message length. Although the difference between the true number of components and that given by the shortest message length is not large, this does serve to illustrate that in this example a larger number of data points are required to `resolve' the true number of components in latent space."

### 2019-09-17 article/mcfa.tex
things

\noindent{}where $\factorloads$ is a $\NumDimensions \times \NumLatentFactors$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata^\mathrm{th}$ data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents^\mathrm{th}$ multivariate
normal distribution, where each $\numcomponents^\mathrm{th}$ component has a $\NumLatentFactors$-dimensional mean and a dense $\NumLatentFactors \times \NumLatentFactors$ covariance matrix.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumLatentFactors \times \NumData$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$. In a sense we
are using latent factor analysis as a form of dimensionality reduction and
simultaneously performing clustering in latent space.

@@

\vspace{1em}

\subsection{Initialisation} \label{sec:initialisation}

@@

\noindent{}Thus the message length to encode $\factorloads$ is given by
\begin{widetext}
\begin{eqnarray}
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad . \label{eq:prior_L}
\end{eqnarray}
%\end{widetext}
Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
%\begin{widetext}
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \label{eq:message-length}
 \end{eqnarray}
\end{widetext}

%\hphantom{asdfasdf\\ asdf \\}
%\vspace{5em}
%\hphantom{asdfasdf\\ asdf \\}
%\vspace{5em}

%\newpage

%\vspace{2em}

@@

	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-mml-contours.pdf} \\
	\end{tabular}
	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-mml-contours.pdf} \\
	\end{tabular}
    \caption{Metrics from our grid search for the toy model for two sample sizes: 
    		 $N = 100,000$ (left) and $N = 5,000$ (right). 
    		 The top panels show the negative log likelihood
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using the generated data in our toy model.
			 The middle panels shows 
			 the BIC (Eq.~\ref{eq:bic}) for those 
			 combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel. A line connects to the true value (black point) to guide the eye.}
    \label{fig:experiment-1-gridsearch}
\end{figure*}
\vspace{-50em}

We recorded the \emph{negative} log likelihood, the BIC, and the message length\footnote{Omitting constant terms such that negative message lengths are allowed.}  for each permutation of $\NumLatentFactors$ and $\NumComponents$.
These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the negative log likelihood decreases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value and message length is found at $\NumLatentFactors = 5$
and $\NumComponents = 10$, identical to the true values.
It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and clustering in the latent space provides a better description of the (generated) data than a Gaussian mixture model without latent factors.
Adding components to the model does improves the log likelihood, even with a single latent factor,
but the addition of just \emph{one latent factor} improves the log likelihood more so than adding
\emph{twenty components}. Not much more can be said for this example because the true data generating process 
is known, but this toy model does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.

@@

\begin{figure*}
	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}

@@

\vspace{2em}

@@

We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances would be untrustworthy. We then took all
stars with a signal-to-noise ratio exceeding 40 in the blue arm (\texttt{snr\_c1 > 40}), and
stars with no erroneous flags in all of the following abundances: Na, Al, Si, K, Ca, Sc, Ti, V, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
There are 1,072 stars that met these criteria.

We executed a grid search for the number of latent factors $\NumLatentFactors$
and the number of components $\NumComponents$ that were preferred by the data.
Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 7$
and $\NumComponents = 5$. 
The results of this grid search are shown in Figure~\ref{fig:exp3-gridsearch},
where we show the negative log likelihood, the BIC, and message length found for each permutation.
The behaviour of the BIC and the message length are very different here, unlike what was observed
in our toy model. Here the BIC behaviour appears similar to the negative log likelihood in that the BIC
prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$). Indeed, if we were to trial higher values of $J$ and $K$ then the negative log likelihood would continue to increase.
The model with five latent factors and three components ($J = 5$, $K = 3$) is found to have the shortest message length, which we take as our preferred model for these data.

Earlier we described how the latent factors we estimate can only be identified up until an orthogonal rotation. If we want to interpret the latent factors estimated from \Galah\ data, then we must specify
some target factor loads such that we can identify which factors are most similar to the yields we expect.
We specified the following target latent factors where:
\begin{itemize}
	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
	\item The third factor load should have non-zero entries in Fe-peak elements Sc, V, Mn, Fe, Ni, Cu, and Zn.
	\item The fourth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Ti.
	\item The fifth factor load should have non-zero entries in the light odd-Z elements Na, Al, and K.
\end{itemize}

@@

Some of these factor load entries may be non-zero because we require the latent factors to
be mutually orthogonal, and not because they truly contribute to the data. To try and
disentangle these possibilities, we calculate the fractional contribution that factor
load makes to the observed abundances relative to other factor loads. We define the
fractional contribution of the $j$th factor load to the $d$th
data dimension as:
\begin{equation}
	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
\end{equation}
The fractional contributions to each element are shown in the right hand side of
Figure~\ref{fig:exp3-factor-loads}. We identify the first factor $\mathbf{L}_1$ 
as being most similar to the r-process, and here it is the dominant contributor to Eu, a typical r-process tracer. Surprisingly we also find that this factor load is a reasonable  contributor to the Fe-peak element Sc. The specific scatter in Sc is 0.03~dex (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the Sc abundances are well-described by this latent factor model.

The second latent factor $\mathbf{L}_2$ here is most representative of the slow neutron capture process (s-process), with dominant contributions to Ba, Y, and La. This factor has some support at other elements, notably K. Most Sc is contributed by $\mathbf{L}_3$, our Fe-peak latent factor. In fact $\mathbf{L}_3$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors. The exception here is Zn, where a near-equal contribution comes from $\mathbf{L}_5$.
The third latent factor $\mathbf{L}_4$ is reasonably well behaved. It is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and surprisingly, Al. It also seems to contribute a non-zero fraction to Eu, with negligible contributions elsewhere.
The specific scatter after accounting for these latent factors is smallest
for Fe (0.01~dex) and largest for K (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}). The typical scatter in most elements is about 0.05~dex.

In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
where the separation between components is arguably best seen in the splitting between $\factorscores_5$ with respect to $\factorscores_2$ or $\factorscores_3$. When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (dark green) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H],
and lower abundance ratios of [V/Fe]. This is consistent with the clustering in latent space.

@@

\vspace{1em}
\section{Discussion} \label{sec:discussion}

@@

We have introduced a data-driven model of nucleosynthesis by incorporating 
latent factors that are common to all stars, and allowing for clustering in the
lower-dimensional latent space. This approach simultaneously allows us to efficiently
tag stars based on their chemical abundances, and to infer the contributions that are
common to all stars (e.g., nucleosynthetic yields). Experiments with generated data
demonstrate that MML is a useful principle for selecting the appropriate number of
latent factors and components. Experiments with \Galah\ data reveal latent factors
that are qualitatively and quantitatively similar to expected nucleosynthetic yields (e.g., products
from the $s$-process, $r$-process, et cetera). Interestingly we find that deviations from
expected yields occur in elements where observations and galactic chemical evolution models
are most discrepant (e.g., K, Sc). While we advise caution in directly interpreting
those latent factors as being nucleosynthetic yields, our
model does provide the first data-driven approach to nucleosynthesis and chemical tagging. We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.\\

\vspace{2em}

%This \emph{Article} was compiled from the \texttt{git} commit \todo{\texttt{HASH}}.
%The analysis, results, and figures presented in this \emph{Article} can be reproduced in full 
%using the following commands in a modern terminal:\vspace{1em}\\
%\texttt{git clone https://github.com/andycasey/mcfa}\\
%\texttt{cd mcfa/}\\
%\todo{\texttt{sh article/build.sh}}

>>>>

\noindent{}where $\factorloads$ is a $\NumDimensions \times \NumLatentFactors$ 
matrix of factor loads that is common to all data points, $\NumLatentFactors$ is
the number of latent factors, and the factor scores 
for the $\numdata^\mathrm{th}$ data point
\begin{equation}
	\factorscores_\numdata \sim \mathcal{N}(\vec\xi_\numcomponents, \vec\Omega_\numcomponents)
\end{equation}
\noindent{}are drawn from\footnote{For clarifying nomenclature across disciplines, the terminology $z \sim \mathcal{N}(0, 1)$ indicates that the $z$ variable \emph{is drawn from} a standard normal distribution.} the $\numcomponents^\mathrm{th}$ multivariate
normal distribution, where each $\numcomponents^\mathrm{th}$ component has a $\NumLatentFactors$-dimensional mean and a dense $\NumLatentFactors \times \NumLatentFactors$ covariance matrix.
The mean vector $\vec\mu$ describes the mean datum in each dimension.
The factor scores for all data points $\factorscores$ is then a 
$\NumLatentFactors \times \NumData$ matrix, where each data point has a partial
association to the components in latent space. 
We assume $\vec{e} \sim \mathcal{N}\left(\vec{0}, \textrm{diag}(\specificvariance)\right)$
is independent of the latent space, and $\specificvariance$ is a
vector of variances in each $\NumDimensions$ abundance dimensions.
In this model each data point can be represented as being drawn
from a mixture of multivariate normal components, except the components
are \emph{clustered in the latent space} $\factorscores$ and projected
into the data space by the factor loads $\factorloads$. In a sense we
are using latent factor analysis as a form of dimensionality reduction and
simultaneously performing clustering in latent space (Figure~\ref{fig:schematic}).

@@

\vspace{1em}
\subsection{Initialisation} \label{sec:initialisation}

@@

\begin{widetext}
\noindent{}Thus the message length to encode $\factorloads$ is given by
\begin{eqnarray}
I(\factorloads)	&=& \frac{1}{2}\textrm{Tr}(\textrm{Cov}(\factorloads\transpose)^{-1}\factorloads\transpose\factorloads) - \frac{1}{2}(\NumDimensions - \NumLatentFactors - 1)\log{|\factorloads\transpose\factorloads|} + \frac{1}{2}\NumDimensions\NumLatentFactors\log{2} + \frac{1}{2}\NumDimensions\log{|\textrm{Cov(\factorloads)}|} - \Gamma\left(\frac{\NumDimensions}{2}\right) \quad . \label{eq:prior_L}
\end{eqnarray}
%\end{widetext}
Combining equations \ref{eq:prior_J}, \ref{eq:prior_K}, \ref{eq:prior_pi}, \ref{eq:prior_xi_omega}, and \ref{eq:prior_L} with equation \ref{eq:mml} leads to the full message length:
%\begin{widetext}
\begin{eqnarray}
	I(\vec\Psi, \vec\data) &=& -\log\likelihood(\vec\data|\vec\Psi)
 +\frac{1}{4}\left(\NumLatentFactors + 4\right)\left(\NumLatentFactors - 1\right)\sum_{\numcomponents=1}^\NumComponents\log\weight_\numcomponents + \left(\NumComponents - \frac{1}{2}\right)\log{\NumData}
 +\frac{1}{2}\NumDimensions\log|\textrm{Cov}\left(\factorloads\transpose\right)| \nonumber \\
  && \cdots -\frac{1}{2}\left(D-J-1\right)\log|\factorloads\transpose\factorloads| + \textrm{Tr}\left(\textrm{Cov}\left(\factorloads\transpose\right)^{-1}\factorloads\transpose\factorloads\right) - \left(\NumLatentFactors + \frac{3}{2}\right)\sum_{\numcomponents=1}^\NumComponents\log|\scorecovs_\numcomponents|  -\log\Gamma\left(\NumComponents\right) - \Gamma\left(\frac{\NumDimensions}{2}\right) \nonumber \\
&& \cdots +\frac{Q}{2}\log\kappa_q +\frac{1}{2}\left[\NumLatentFactors(\NumDimensions + 2) + \NumComponents(2-\NumData)\right]\log{2}  \quad . \label{eq:message-length}
 \end{eqnarray}
 \newpage
\end{widetext}

@@

	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-1-gridsearch-mml-contours.pdf} \\
	\end{tabular}
	\begin{tabular}[b]{@{}p{0.45\textwidth}@{}}
		\centering\includegraphics[width=\linewidth]{experiments/eval-figs/eval-2-gridsearch-mml-contours.pdf} \\
	\end{tabular}
    \caption{Metrics from our grid search for the toy model for two sample sizes: 
    		 $N = 100,000$ (left) and $N = 5,000$ (right). 
    		 The top panels show the negative log likelihood
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors 
			 $\NumLatentFactors$ and number of clusters 
			 $\NumComponents$ using the generated data in our toy model.
			 The middle panels shows 
			 the BIC (Eq.~\ref{eq:bic}) for those 
			 combinations, and the lower panel shows the 
			 message length. The white marker indicates the
			 lowest value in each panel. A line connects to the true value (black point) to guide the eye.}
    \label{fig:experiment-1-gridsearch}
    \vspace{-10em}
\end{figure*}

We recorded the \emph{negative} log likelihood, the BIC, and the message length\footnote{Omitting constant terms such that negative message lengths are allowed.}  for each permutation of $\NumLatentFactors$ and $\NumComponents$.
These metrics are shown in Figure~\ref{fig:experiment-1-gridsearch}.
Unsurprisingly the negative log likelihood decreases with increasing numbers of latent
factors $\NumLatentFactors$ and increasing numbers of components $\NumComponents$.
The lowest BIC value and message length is found at $\NumLatentFactors = 5$
and $\NumComponents = 10$, identical to the true values.

We repeated this toy model experiment using a smaller sample size ($N = 5,000$) to be more representative of the sample sizes in later \Galah\ experiments (Section~\ref{sec:exp4}). The results of the grid search are also shown in Figure~\ref{fig:experiment-1-gridsearch}. Here BIC estimates the true number of latent factors correctly, but tends to underestimate the true number of clusters, more so than the message length. Although the difference between the true number of components and that given by the shortest message length is not large, this does serve to illustrate that in this example a larger number of data points are required to `resolve' the true number of components in latent space.

It is clear from Figure~\ref{fig:experiment-1-gridsearch} that a combination of latent factors
and clustering in the latent space provides a better description of the (generated) data than a Gaussian mixture model without latent factors.
Adding components to the model does improves the log likelihood, even with a single latent factor,
but the addition of just \emph{one latent factor} improves the log likelihood more so than adding
\emph{twenty components}. Not much more can be said for this example because the true data generating process 
is known, but this toy model does illustrate how 
clustering in high dimensional data can be better described by latent factors with 
clustering in the lower dimensional latent space.

@@

@@

@@

We first selected stars with \texttt{flag\_cannon = 0} to exclude
stars where there is reason to suspect that the stellar parameters
(e.g., $\teff$, $\logg$) are unreliable, and as a result the 
detailed chemical abundances would be untrustworthy. We then took all
stars with a signal-to-noise ratio exceeding 30 in the blue arm (\texttt{snr\_c1 > 30}), and
stars with no erroneous flags in all of the following abundances: 
Mg, Na, Al, Si, K, Ca, Sc, Ti, Mn, Fe, Ni, Cu, Zn, Y, Ba, La, and Eu.
These elements were chosen because they trace multiple nucleosynthetic pathways, and they are more commonly reported in the \Galah\ data release, allowing for a larger number of stars with a complete abundance inventory.
There are 2,566 stars that met these criteria.

We executed a grid search for the number of latent factors $\NumLatentFactors$
and the number of components $\NumComponents$ that were preferred by the data.
Starting with $\NumLatentFactors = 1$ and $\NumComponents = 1$, we trialled each
permutation of $\NumLatentFactors$ and $\NumComponents$ up until $\NumLatentFactors = 7$
and $\NumComponents = 5$. 
The results of this grid search are shown in Figure~\ref{fig:exp3-gridsearch},
where we show the negative log likelihood, the BIC, and message length found for each permutation.
The behaviour of the BIC and the message length are very different here, unlike what was observed
in our toy model. Here the BIC behaviour appears similar to the negative log likelihood in that the BIC
prefers higher components and latent factors than the extent of the grid (e.g., $J > 7$ and $K > 5$). Indeed, if we were to trial higher values of $J$ and $K$ then the negative log likelihood would continue to increase.
The model with five latent factors and three components ($J = 6$, $K = 3$) is found to have the shortest message length, which we take as our preferred model for these data.

Earlier we described how the latent factors we estimate can only be identified up until an orthogonal rotation. If we want to interpret the latent factors estimated from \Galah\ data, then we must specify
some target factor loads such that we can identify which factors are most similar to the yields we expect.
We specified the following target latent factors where:
\begin{itemize}
	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
	\item The third factor load should have non-zero entries in Fe-peak elements V, Mn, Fe, Ni, Zn, and Ti.
	\item The fourth factor load should have non-zero entries in the odd-Z elements Na, Al, K, Sc, and Cu.
	\item The fifth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Mg.
%	\item The first factor load should have non-zero entries in Eu and La (e.g., the $r$-process).
%	\item The second factor load should have non-zero entries in Ba, Y, and La (e.g., the $s$-process).
%	\item The third factor load should have non-zero entries in Fe-peak elements Sc, V, Mn, Fe, Ni, Cu, and Zn.
%	\item The fourth factor load should have non-zero entries in the $\alpha$-element tracers Si, Ca, and Ti.
%	\item The fifth factor load should have non-zero entries in the light odd-Z elements Na, Al, and K.
\end{itemize}

\begin{figure*}[t!]
	%\includegraphics[width=\textwidth]{experiments/exp1-compare-all.pdf}
	\includegraphics[width=\textwidth]{experiments/eval-figs/eval-1-compare-all.pdf}
	\caption{The estimated factor loads $\factorloads$ (left), factor scores $\factorscores$ (middle),
    		 and specific variances $\specificvariance$ (right) compared to the 
		 	 true data generating values
		 	 for Experiment~1 (Section~\ref{sec:exp-1}). The agreement is excellent.}
    \label{fig:exp1-compare}
\end{figure*}

@@

Some of these factor load entries may be non-zero because we require the latent factors to
be mutually orthogonal, and not because they truly contribute to the data. To try and
disentangle these possibilities, we calculate the fractional contribution that factor
load makes to the observed abundances relative to other factor loads. We define the
fractional contribution of the $j$th factor load to the $d$th
data dimension as:
\begin{equation}
	\mathbf{C}_\textrm{d,j} = \frac{\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|}{\sum^J\sum^{N}|\factorloads_{j,d}\factorscores_{n,j}|} \quad .
\end{equation}
The fractional contributions to each element are shown in the right hand side of
Figure~\ref{fig:exp3-factor-loads}. We identify the first factor $\mathbf{L}_1$ 
as being most similar to the r-process, and here it is the dominant contributor to Eu, a typical r-process tracer. Surprisingly we also find that this factor load is a reasonable contributor to the odd-Z element Sc. The specific scatter in Sc is 0.03~dex (Figure~\ref{fig:exp3-specific-scatter}), suggesting that the Sc abundances are well-described by this latent factor model.

The second latent factor $\mathbf{L}_2$ here is most representative of the slow neutron capture process (s-process), with dominant contributions to Ba, and Y. This factor has some support at other elements, notably K. $\mathbf{L}_3$ is the primary contributor to nearly all Fe-peak elements, with close to negligible contributions from other factors. The exception here is Cu, where a near-equal contribution comes from $\mathbf{L}_4$.
The fifth latent factor $\mathbf{L}_5$ is the dominant contributor to the $\alpha$-element tracers Si, Ca, and Ti, and surprisingly, Al. 
The specific scatter after accounting for these latent factors is smallest
for Fe (0.01~dex) and largest for K (0.13~dex; Figure~\ref{fig:exp3-specific-scatter}). The typical scatter in most elements is about 0.05~dex.

In Figure~\ref{fig:exp3-latent-space} we show the inferred clustering in latent space,
where the separation between components is arguably best seen in the splitting between $\factorscores_6$ with respect to $\factorscores_2$ or $\factorscores_3$. When projected to data space (Figure~\ref{fig:exp3-data-space}) the third component (light green) is seen to have relatively higher abundance ratios of [K,Ba,Zn/Fe] at a given [Fe/H]. This is consistent with the clustering in latent space.

@@

\section{Discussion} \label{sec:discussion}

@@

We have introduced a data-driven model of nucleosynthesis by incorporating 
latent factors that are common to all stars, and allowing for clustering in the
lower-dimensional latent space. This approach simultaneously allows us to efficiently
tag stars based on their chemical abundances, and to infer the contributions that are
common to all stars (e.g., nucleosynthetic yields). Experiments with generated data
demonstrate that MML is a useful principle for selecting the appropriate number of
latent factors and components. Experiments with \Galah\ data reveal latent factors
that are qualitatively and quantitatively similar to expected nucleosynthetic yields (e.g., products
from the $s$-process, $r$-process, et cetera). Interestingly we find that deviations from
expected yields occur in elements where observations and galactic chemical evolution models
are most discrepant (e.g., K, Sc). While we advise caution in directly interpreting
those latent factors as being nucleosynthetic yields, our
model does provide the first data-driven approach to nucleosynthesis and chemical tagging. We advocate that more data, and the inclusion of weakly informative priors -- joint probabilities using astrometry and a simplified model of the Milky Way -- would help in realising the full potential of chemical tagging.
